---
title: "Package Installation"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(eval=FALSE)
```

We are going to install [Seurat](https://satijalab.org/seurat/). Be sure to check out their [vignettes](https://satijalab.org/seurat/articles/)!

First, ensure you're running R `4.1.3`.

The next code block only needs to be run **once**. If asked to update packages, answer 'none'. We take care of it later.

```{r}
installed_pkgs <- installed.packages()

if (! "markdown" %in% installed_pkgs) install.packages("markdown") 

if (! "remotes" %in% installed_pkgs) install.packages("remotes") 

if (! "BiocManager" %in% installed_pkgs) install.packages("BiocManager")
```

With `BiocManager`, we can install Bioconductor `3.14`.

```{r}
BiocManager::install(version='3.14')
```

> You can ignore the message:
> 
> ```
> Bioconductor version 3.14 (BiocManager
>   1.30.20), R 4.1.3 (2022-03-10)
> Installation paths not writeable, unable to
>   update packages
>   path: /opt/R/4.1.3/lib/R/library
>   packages:
>     boot, class, cluster, codetools, foreign,
>     MASS, Matrix, mgcv, nlme, nnet, rpart,
>     spatial, survival
> ```

R packages bundle data with them, usually for testing purposes. In the case of `Seurat`, there is a package (`SeuratData`) specifically designed to download some datasets. You may use `SeuratData::AvailableData()` to get a table of all these 'educational' datasets. Before proceeding with the installation of other packages, let's get a dataset:

```{r}
if (! "SeuratData" %in% installed_pkgs) remotes::install_github('satijalab/seurat-data')
if (! "ifnb.SeuratData" %in% installed_pkgs) SeuratData::InstallData("infb")
if (! "pbmc3k.SeuratData" %in% installed_pkgs) SeuratData::InstallData("pbmc3k")
```

We are going to use another package available on github:

```{r}
if (! "kBET" %in% installed_pkgs) remotes::install_github('theislab/kBET')
```

The next code block defines `my_load_namespaces` function that takes two named arguments: `packagesToLoad` and `coreLibraries`. Both are vectors of strings, corresponding to packages that need to be installed. Then, the function loads each of the packages specified in `packagesToLoad` vector, through the use of the `require` base R function. The `coreLibraries` are not explicitly loaded, but required anyway (e.g. dependencies). If the execution of the whole function is alright, it will simply return `TRUE`.

```{r}
my_load_namespaces <- function(packagesToLoad=c(), coreLibraries=c()) {
  
  installed_pkgs <- installed.packages()
  
  if (! "BiocManager" %in% installed_pkgs) install.packages("BiocManager")
  
  lapply(c(coreLibraries, packagesToLoad),
         function(x) if (! x %in% installed_pkgs) BiocManager::install(x, ask = FALSE))
  
  lapply(packagesToLoad, function(pkg) {
    suppressPackageStartupMessages(require(pkg, character.only = TRUE))
  })
  
  TRUE
}
```

Let's use it! The whole process is going to take a couple of minutes...

```{r}
my_load_namespaces(
  packagesToLoad = c("openxlsx", "readr", "magrittr", "dplyr", "RColorBrewer", "ggplot2", "patchwork", "Seurat", "future","glmGamPoi"),
  coreLibraries = c("knitr", "rmarkdown", "formatR", "DT", "reticulate", "sessioninfo", "limma", "DESeq2", "uwot", "future.apply", "metap", "enrichR", "shiny", "SingleCellExperiment", "remotes", "bookdown")
)
```

Sometimes, there are errors while installing packages in bulk that are easily solved by re-iterating the command. These error messages are difficult to track, since we get so much output from the ongoing process in bulk, and also because of all the dependencies. This is specially true for first-comers, because their library of packages is empty. Anyhow, just re-run the previous code block 2-3 times until its only output is: `TRUE`.

Finally, we want our installation to be coherent:

```{r}
stopifnot(BiocManager::valid())
```

If the above code block fails, is because according to BiocManager our current state is not valid (so, it didn't return `TRUE` to `stopifnot`). To correct this, you should run the `BiocManager::install()` command as it's stated by `BiocManager::valid()`.


# Download preprocessed datasets from zenodo

For the later part of the course, we have preprocessed a couple of large datasets and would like you to download them from zenodo:

```{r zenodo}
if (!file.exists("datasets/preprocessed_rds/panc_sub.RDS")){
utils::download.file("https://zenodo.org/record/7848230/files/panc_sub.RDS?download=1",destfile="datasets/preprocessed_rds/panc_sub.RDS",method="wget")}
if (!file.exists("datasets/preprocessed_rds/ref.labels.RDS")){
utils::download.file("https://zenodo.org/record/7845337/files/ref.labels.RDS?download=1",destfile="datasets/preprocessed_rds/ref.labels.RDS",method="wget")}
if (!file.exists("datasets/preprocessed_rds/pbmc_ref.seurat.RDS")){
utils::download.file("https://zenodo.org/record/7845337/files/pbmc_ref.seurat.RDS?download=1",destfile="datasets/preprocessed_rds/pbmc_ref.seurat.RDS",method="wget")}

```