---
title: "Data Visualization"
date: "`r format(Sys.time(), '%d %B, %Y')`"

output:
  bookdown::html_document2:
    theme: spacelab
    highlight: monochrome
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 2
    df_print: paged

params:
  Nthreads:
    label: 'Number of Threads:'
    value: 8
    input: slider
    min: 4
    max: 96
  Nmemgb:
    label: 'Gigabytes of Memory per Thread:'
    value: 8
    input: slider
    min: 4
    max: 16
  fork:
    label: 'Fork processes (not GUI)'
    value: FALSE
    input: checkbox
  seed:
    label: 'Random seed:'
    value: 8211673
  dimensionality:
    label: 'Dimensionality of single-cell dataset:'
    value: 10
  cluster_resolution:
    label: 'Parameter for clustering algorithm:'
    value: 0.5
---


# Setup

```{r}
timestamp()
```

```{r}
getwd()
```


```{r}
print(params)
```

```{r}
my_load_namespaces <- function(packagesToLoad=c(), coreLibraries=c()) {
  installed_pkgs <- installed.packages()
  if (! "BiocManager" %in% installed_pkgs) install.packages("BiocManager")
  lapply(c(coreLibraries, packagesToLoad),
         function(x) if (! x %in% installed_pkgs) BiocManager::install(x, ask = FALSE))
  lapply(packagesToLoad, function(pkg) {
    suppressPackageStartupMessages(require(pkg, character.only = TRUE))
  })
  TRUE
}

my_load_namespaces(
  packagesToLoad = c("openxlsx", "readr", "magrittr", "dplyr", "RColorBrewer", "ggplot2", "patchwork", "Seurat"),
  coreLibraries = c("knitr", "rmarkdown", "formatR", "DT", "reticulate", "sessioninfo", "limma", "DESeq2", "enrichR", "future", "uwot", "future.apply", "metap", "shiny", "SingleCellExperiment", "remotes", "bookdown", "SeuratData", "scCustomize", "Nebulosa")
)
```

```{r}
stopifnot(suppressMessages(BiocManager::valid()))

set.seed(params$seed)
knitr::opts_chunk$set(echo = TRUE)

options(parallelly.fork.enable = params$fork,
        future.globals.maxSize = params$Nmemgb * 1024^2 * 1000)

plan("multicore", workers = params$Nthreads)

myFlappy <- function(x, ...) {
  future.apply::future_lapply(x, ..., future.seed = TRUE)
}
```

## Load Data

We'll be working with the data from Day 1, `14_FirstSteps.Rmd`, let's quickly re-load and re-process again:

```{r}
pbmc <- Read10X(data.dir = "../datasets/filtered_gene_bc_matrices/hg19/") %>%
  CreateSeuratObject(counts = ., project = "pbmc3k", min.cells = 3, min.features = 200)

pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)

pbmc <- NormalizeData(pbmc, verbose = FALSE)

pbmc <- FindVariableFeatures(pbmc, verbose = FALSE)

pbmc <- ScaleData(pbmc, features = rownames(pbmc), verbose = FALSE)

pbmc <- RunPCA(pbmc, features = VariableFeatures(pbmc), verbose = FALSE)

pbmc <- FindNeighbors(pbmc, dims = seq_len(params$dimensionality), verbose = FALSE)

pbmc <- FindClusters(pbmc, resolution = params$cluster_resolution, verbose = FALSE)

pbmc <- RunUMAP(pbmc, dims = seq_len(params$dimensionality), verbose = FALSE)

new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B", "CD8 T", "FCGR3A+ Mono", "NK", "DC", "Platelet")
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)
rm(new.cluster.ids)
```

# scCustomize

https://satijalab.org/seurat/articles/visualization_vignette.html


# End

HOMEWORK: Install python-umap in a VENV, and re-run the algorithm using this implementation rather than the UWOT. Are there any differences in the results? If so, did you reset the seed in between runs? What about the time it takes? You can measure with `microbenchmark` package.
