---
title: "Shiny brushedPlot"
date: "`r format(Sys.time(), '%d %B, %Y')`"

output:
  bookdown::html_document2:
    number_sections: false
    theme: spacelab
    highlight: monochrome
    code_folding: show
    toc: true
    toc_float: true
    toc_depth: 2
    df_print: paged
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", eval=FALSE, format=TRUE)
```

Consider these extra notes after learning Seurat.

## Load Data

We'll be working with the data from Day 1 ("First steps"), let's quickly re-load and re-process again:

```{r}
pbmc <- Read10X(data.dir = "./datasets/filtered_gene_bc_matrices/hg19/") %>%
  CreateSeuratObject(counts = ., project = "pbmc3k", min.cells = 3, min.features = 200)

pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
pbmc <- NormalizeData(pbmc, verbose = FALSE)
pbmc <- FindVariableFeatures(pbmc, verbose = FALSE)
pbmc <- ScaleData(pbmc, features = rownames(pbmc), verbose = FALSE)
pbmc <- RunPCA(pbmc, features = VariableFeatures(pbmc), verbose = FALSE)
pbmc <- FindNeighbors(pbmc, dims = seq_len(params$dimensionality), verbose = FALSE)
pbmc <- FindClusters(pbmc, resolution = params$cluster_resolution, verbose = FALSE)
pbmc <- RunUMAP(pbmc, dims = seq_len(params$dimensionality), verbose = FALSE)

new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B", "CD8 T", "FCGR3A+ Mono", "NK", "DC", "Platelet")
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)
rm(new.cluster.ids)
```

## Cell-Cycle Scoring

We didn't try to mitigate the effects of cell cycle heterogeneity in the pbmc3k dataset. Follow the [tutorial](https://satijalab.org/seurat/articles/cell_cycle_vignette.html) if the need arises when working with your own dataset.

We assign scores in the `CellCycleScoring()` function, which stores S and G2/M scores in object meta data, along with the predicted classification of each cell in either G2M, S or G1 phase. This function uses some custom lists of genes that are already bundled, it's mainly a wrapper around `AddModuleScore()`.

```{r}
# Better safe than sorry. Avoid false-positives.
standardize.gene.name <- function(gene_name, sp) {
  if (sp == "mouse") {
    x <- tools::toTitleCase(tolower(gene_name))
  } else if (sp == "human") {
    x <- toupper(gene_name)
  }
  x
} 

pbmc %<>% CellCycleScoring(s.features = standardize.gene.name(cc.genes$s.genes, sp="human"),
                           g2m.features = standardize.gene.name(cc.genes$g2m.genes, sp="human"),
                           nbin = 20)

# To allow using 'alternate' cell-cycle regress-out @ PCA
pbmc$CC.Difference <- pbmc$S.Score - pbmc$G2M.Score
```

To understand how these scores are calculated, take a look at the [original research paper](https://doi.org/10.1126/science.aad0501). Here's an excerpt:

> The top 100 MITF-correlated genes across the entire set of malignant cells were defined as the MITF program, and their average relative expression as the MITF-program cell score. The average expression of the top 100 genes that negatively correlate with the MITF program scores were defined as the AXL program and used to define AXL program cell score. To decrease the effect that the quality and complexity of each cell’s data might have on its MITF/AXL scores we defined control gene-sets and their average relative expression as control scores, for both the MITF and AXL programs. These control cell scores were subtracted from the respective MITF/AXL cell scores. The control gene-sets were defined by first binning all analyzed genes into 25 bins of aggregate expression levels and then, for each gene in the MITF/AXL gene-set, randomly selecting 100 genes from the same expression bin as that gene. In this way, a control gene-sets have a comparable distribution of expression levels to that of the MITF/AXL gene-set and the control gene set is 100-fold larger, such that its average expression is analogous to averaging over 100 randomly-selected gene-sets of the same size as the MITF/AXL gene-set.

## Shiny

<!-- https://shiny.rstudio.com/gallery/shiny-wysiwyg.html -->

Check out `27_brushedPlot.R`.

## openXLSX

The following `save_XLSX()` function is to save all your DE tests results in a XLSX file (e.g. for sharing with a colleague that's not so proficient in R-language).

```{r, eval=FALSE}
#' Export Data Frames to a Microsoft Excel File
#'
#' Creates a workbook with different sheets, each corresponding to a data frame that was passed in the first argument. This would also split by clusters to different sheets, if this column is found.
#' @param dfl A list of data frames, like the ones returned by your DE testing.
#' @param filename A file.path with the extension XLSX.
#' @return No return, just the side-effect of writing data to disk in the appropriate format to be shared with colleagues.
#' @examples
#' save_XLSX(list(PerCluster=markers.between.clusters, WithinClusters=markers.within.clusters), file.path(getwd(), paste0("../results/results_", format(Sys.time(), '%d_%B__%H-%M-%S_%Y'), ".xlsx")))
save_XLSX <- function(dfl, filename) {
  wb <- createWorkbook()
  options(openxlsx.borderColour = "#4F80BD")
  options(openxlsx.borderStyle = "thin")
  modifyBaseFont(wb, fontSize = 14, fontName = "Arial Narrow")
  for (df_id in names(dfl)) {
    if ("cluster" %in% colnames(dfl[[df_id]])) {
      for (cluster_n in seq_len(max(dfl[[df_id]]$cluster))) {
        sheet_name <- paste(df_id, cluster_n, sep="_")
        my_data <- dfl[[df_id]][dfl[[df_id]]$cluster == cluster_n, ]
        my_data$gene <- sub(
          '\\(.*\\)\\.[0-9]', '\1',
          my_data$gene
          )
        my_data$cluster <- NULL
        addWorksheet(wb, sheetName = sheet_name, gridLines = FALSE)
        writeDataTable(wb, sheet = sheet_name,
                       x = my_data,
                       colNames = TRUE, rowNames = FALSE,
                       tableStyle = "TableStyleLight9")
      }
    } else {
      addWorksheet(wb, sheetName = df_id, gridLines = FALSE)
      writeDataTable(wb, sheet = df_id,
                     x = dfl[[df_id]],
                     colNames = TRUE, rowNames = FALSE,
                     tableStyle = "TableStyleLight9")
    }
  }
  saveWorkbook(wb, filename, overwrite = TRUE)
}
```

## Viz: scCustomize

<!-- https://satijalab.org/seurat/articles/visualization_vignette.html#interactive-plotting-features

https://samuel-marsh.github.io/scCustomize/articles/Gene_Expression_Plotting.html#shuffle-points

https://samuel-marsh.github.io/scCustomize/reference/index.html -->

This library provides with extra arguments, functions, and behaviors over the default Seurat. This isn't adding boilerplate code, but on the contrary. It will also provide us with an interface to enforce common plotting requirements.

Using the customized plotting functions keep us away from reinventing the wheel. For example: Extracting the Top N=10 (or whatever `N`) genes per identity after running `Seurat::FindAllMarkers()` is very common and `scCustomize` provides `Extract_Top_Markers()` function to this end. This is extremely helpful to avoid introducing bugs.

### QC

#### Threshold lines

```{r}
pbmc <- Read10X(data.dir = "./datasets/filtered_gene_bc_matrices/hg19/") %>%
  CreateSeuratObject(counts = ., project = "pbmc3k", min.cells = 3, min.features = 200)

pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
```


```{r, fig.asp=1.5}
QC_Plots_Genes(pbmc, high_cutoff = 2500, low_cutoff = 200)
```

#### Volcanoes

```{r}
pbmc <- FindVariableFeatures(pbmc)
```

Remember how we extracted feature names into a vector, `top10`, and then used Seurat's `LabelPoints()`? Don't you? No worries:

```{r}
VariableFeaturePlot_scCustom(pbmc, num_features = 10, repel = TRUE, y_axis_log = TRUE) + NoLegend()
```


#### UMI vs Features

Remember our Scatter Plot? and the Percent of Mitochondrial genes Violinn Plot?

```{r, fig.asp=1.2}
QC_Plots_Mito(pbmc, high_cutoff = 15, mito_name = "percent.mt")
```

```{r, fig.asp=1.2}
QC_Plot_UMIvsGene(pbmc, low_cutoff_gene = 200, high_cutoff_gene = 2500, meta_gradient_name = "percent.mt")
```

```{r}
QC_Plot_GenevsFeature(pbmc, "percent.mt", low_cutoff_gene = 200, high_cutoff_gene = 2500, high_cutoff_feature = 15) + NoLegend()
```

> ⌨🔥 Exercise: Try plotting with QC_Plot_UMIvsFeature()


```{r}
# Idents(pbmc) <- "seurat_clusters"
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
pbmc <- NormalizeData(pbmc, verbose = FALSE)
pbmc <- FindVariableFeatures(pbmc, verbose = FALSE)
pbmc <- ScaleData(pbmc, features = rownames(pbmc), verbose = FALSE)
pbmc <- RunPCA(pbmc, features = VariableFeatures(pbmc), verbose = FALSE)
pbmc <- FindNeighbors(pbmc, dims = seq_len(params$dimensionality), verbose = FALSE)
pbmc <- FindClusters(pbmc, resolution = params$cluster_resolution, verbose = FALSE)
pbmc <- RunUMAP(pbmc, dims = seq_len(params$dimensionality), verbose = FALSE)

new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B", "CD8 T", "FCGR3A+ Mono", "NK", "DC", "Platelet")
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)
rm(new.cluster.ids)
```

With the same data, after the pipeline is applied, we can do some new plots. Some, are new takes on the same pair of axis, but with the novel advantage of having cell `Identities` set to the current `seurat_clusters`.

> ⌨🔥 Exercise: Plot "Mito Gene % per Cell/Nucleus" for each cell type. Hint: `group.by` argument from previously used function.

Note: `scCustomize::Add_Mito_Ribo_Seurat()` is the built-in function to add percentage of mitochondrial and ribosomal genes in two different metadata columns. This could be prefered, and would change our arguments having the string `percent.mt` to the new string value: `percent_mito`.

### UMAP

```{r, fig.asp=1.1}
DimPlot_scCustom(pbmc, split.by = "seurat_clusters", split_seurat = FALSE, num_columns = 3, pt.size = 0.5, repel = TRUE) + NoLegend()
```

```{r}
FeaturePlot_scCustom(pbmc, features = "CD8A", alpha_exp = 0.5)  # CD8+ T cells
```

```{r}
Plot_Density_Joint_Only(pbmc, features = c("GNLY", "NKG7"))  # NK cells
```

Note: There's `test.use = "ROC"` for DE analyses, this returns the 'classification power' for any individual marker (ranging from 0 - random, to 1 - perfect).

On default `Seurat::FeaturePlot()`, we had either of these two:

```{r, fig.asp=0.5}
FeaturePlot(pbmc, features = c("GNLY", "NKG7"))
```

```{r, fig.asp=0.25}
FeaturePlot(pbmc, features = c("GNLY", "NKG7"), blend = TRUE)
```

<!-- markers taken from literature/ here: https://satijalab.org/seurat/articles/pbmc3k_tutorial.html#assigning-cell-type-identity-to-clusters -->

### DE

```{r}
myFlappy(features, function(i) Nebulosa::plot_density(pbmc, i))
```


> ⌨🔥 Exercise(s):

> 1. Install python's umap in a VENV, and re-run the algorithm using this implementation rather than R-lang's UWOT. Hint: either slide [\#32](http://wiki.immunbio.mpg.de/wiki/images/0/00/02.pdf) of our past Adv. Linux course, or relevant answer on our [FAQ](http://wiki.immunbio.mpg.de/wiki/index.php/Rstudio#Can_I_use_Python_environments_on_RStudio_IDE.3F).


> 2. Are there any differences in the results? If so, did you reset the seed between runs? What about the time it takes? You can measure with `microbenchmark` package.