---
title: "Dataset Challenge"
author: "Katarzyna Sikora"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: spacelab
    highlight: pygments
    code_folding: show
    toc: true
    toc_float: true
    toc_depth: 2
    df_print: paged
---

## Setup

```{r, setup, include=FALSE}
# .libPaths("/rstudio/sikora/rstudio/R/workbench-library/4.1")
.libPaths()
knitr::opts_chunk$set(fig.width=6, fig.height=4,message=FALSE,warning=FALSE,tidy=TRUE, tidy.opts=list(width.cutoff=70)) 
```

## Problem Description

We're going to list datasets bundled with the scRNA Bioconductor package. Pick a dataset of your liking and assign it to an object.
Remember to downsample to a low-k number of cells to limit computational resources for the duration of the class!
Which class is the object ?

Pick a task you'd like to work on, and good luck!


```{r get_data}
library(scRNAseq)
listDatasets()
```

## SessionInfo

```{r sessionInfo}
sessionInfo()
```

## Citations

Risso D, Cole M (2023). scRNAseq: Collection of Public Single-Cell RNA-Seq Datasets. R package version 2.14.0. 
