---
title: "Misc"
author: "Thomas Manke"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: spacelab
    highlight: pygments
    code_folding: show
    toc: true
    toc_float: true
    toc_depth: 2
    df_print: paged
---

```{r echo=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

> This markdown will not run on its own !!!

This is a collection of code snippets to go with the RSeurat course.
They all assume that the respective Seurat objects are loaded.

# Top expressed genes
```{r top_genes}
top20 <- pbmc.data %>% rowSums() %>% sort() %>% tail(20) %>% rev() %>% names()
mat <- pbmc.data[top20,] %>% as.matrix() %>% t()

mat %>%
  as.data.frame() %>%
  gather("name","count") %>% 
  ggplot(aes(x=name,y=count, col=name)) +
  geom_point() + 
  coord_flip() +
  NoLegend()

RidgePlot(pbmc, features = top20[1:6], col=3) + NoLegend()
```

Messages:

- ribosomes are most abundant: healthy cells have plenty of ribosomes
- still the counts per cell are small (<100) compared to bulk RNA
- not seen here: distributions over cells can be very complex and multi-modal (diff. cell types)

## Ribosomes and Mitochondria
```{r}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
pbmc[["percent.rp"]] <- PercentageFeatureSet(pbmc, pattern = "^RP[SL]")
FeatureScatter(pbmc, feature1 = "percent.rp", feature2 = "percent.mt")

pbmc@meta.data %>%
  ggplot(aes(x=percent.rp, y=percent.mt, col=log(nCount_RNA+1))) +
  geom_point() +
  scale_color_gradient(low="lightblue", high="red")
```

# Cell-cell correlation
```{r}
# two cell with large count
top2 <- pbmc.data %>% colSums() %>% sort() %>% rev() %>% tail(2) %>% names()
pbmc.data[,top2] %>% as.data.frame() %>%
  ggplot(aes(.data[[top2[1]]], .data[[top2[2]]])) +
  geom_point()
  
# two cells from same cluster1
(top2_1 <- pbmc@meta.data %>% filter(seurat_clusters==1) %>% arrange(-nCount_RNA) %>% head(2) %>% rownames())
lookup <- c(cell1 = "AACCTACTGTGAGG-1", cell2 = "ACCCAGCTGTTAGC-1")

pbmc.data[,top2_1] %>% as.data.frame() %>%
  rename(all_of(lookup)) %>%
  ggplot(aes(x=cell1,y=cell2)) +
  geom_point()
```



# Data Inspection

... from overall metadata info
```{r}
panc_sub$nCount_RNA %>% hist(2000, xlim=c(0,50000))               # counts per cell distribution
panc_sub$nCount_RNA %>% sort() %>% rev() %>% log10() %>% plot()   # rank plot  
```

```{r}
# illustrate count differences
counts <- GetAssayData(panc_sub, slot = "counts")
hist(counts["ACTB",], 5000, xlim=c(0,200))  # highly expressed, most cells + dropouts
hist(counts["CD4",], 5000, ylim=c(0,20))    # weak expressed, few cell types
```

# Batch testing
Here I wanted to illustrate the effect of batch effect "removal" (with Seurat:CCA)
on the correlation of the batch variable with the leading PC components
(c.f lecture rmd/30..)

```{r R2}
b <- panc_sub$dataset %>% as.factor() # batch variable

myR2 <- function(y){ summary(lm(y ~ b))$r.squared }


# get PCA embeddings for the initial data "panc_sub" and batch controlled integration "panc_combined"
R2 <- data.frame(
  merged= apply(Embeddings(panc_sub),2,myR2)[1:30],          # simple merge - no batch correction
  integrated = apply(Embeddings(panc.combined),2,myR2)       # integrated data - with batch correction
)

R2 %>%
  rownames_to_column(var = "PC") %>%
  mutate(PC=as.factor(as.integer(gsub("PC_","",PC)))) %>%
  gather(status, "R2", -PC ) %>%
  ggplot(aes(x=PC,y=R2, fill=status)) +
  geom_bar(stat = "identity", position = 'dodge') +
  theme(
    axis.text = element_text(size=8)
  )
```
### Multiple Multivariate Regression
```{r multiple_reg}
pca <- Embeddings(panc_sub)      # score matrix
b <- as.factor(panc_sub$dataset) # batch variable

#fit1 <- lm(pca[,1] ~ b )    # regression of PC_1 ~ b
fit <-  lm(pca ~ b )        # multiple regression all PC_i ~ b

# fit is a list of fits and summary a list of summaries
ls <- summary(fit)
# return r.squared from each summary s
r2 <- lapply(ls, function(s) { s$r.squared}) %>% unlist()
# trim names for easier plotting
names(r2) <- sub('Response ','', names(r2))

barplot(r2, las=2, cex.axis = 0.6)

# canonical correlation coefficients are only defined for numeric X,Y
# forcing b --> 0,1 is poor style

#cancor(pca[,1], as.numeric(b))$cor
cancor(pca, as.numeric(b))$cor
```

### Canonical Correlation
```{r canonical_corr}
pca1 <- Embeddings(panc_sub)        # score matrix
pca2 <- Embeddings(panc.combined)

x <- panc_sub$nCount_RNA         # batch variable: sequencing depth
cancor(pca1, x)$cor
cancor(pca2, x)$cor
```

## Variance explained 
Identify any variable in metadata that has an impact on variance (biological or technical)
```{r}
myR2 <- function(y){ summary(lm(y ~ b))$r.squared }

pca <- Embeddings(panc_sub)[,1:30]

vars <- panc_sub@meta.data %>% colnames() # list of all vars to consider

df <- data.frame()

for (v in vars) {

  b <- panc_sub@meta.data %>% pull(v) 
  if (!is.numeric(b)) { 
      b <- as.factor(b) 
      if (nlevels(b) < 2) { next }
  } 

  if (v == "tech") { next }  # redundant with "dataset"
  print(v, str(b))
  
  r2s = apply(pca,2,myR2)
  
  # convert to data frame
  add = data.frame(R2=r2s, v=v) %>%  
    rownames_to_column(var = "PC") %>%
    mutate(PC=as.factor(as.integer(gsub("PC_","",PC))))
  
  df = rbind(df, add) # append
}
 
ggplot(df, aes(x=PC,y=R2, fill=v)) +
  geom_bar(stat = "identity", position = 'dodge') +
  facet_wrap(~v) + 
  ggtitle("Test") +
  theme(
    axis.text = element_text(size=8)
    )

```
## Distribution in UMAP

How different metadata is distributed in UMAP
```{r}
vars <- panc_sub@meta.data %>% colnames()                    # list of all vars to consider
num_vars <- panc_sub@meta.data %>% select(where(is.numeric)) %>% colnames()
cat_vars <- c("orig.ident","celltype","dataset")

panc_sub@meta.data %>% select_if(negate(is.numeric)) %>% colnames()

FeaturePlot(panc_sub, features = num_vars)

DimPlot(panc_sub, group.by=cat_vars)  & NoLegend()

```


# Compare scree plots before and after integration

Somewhere I read that we could/should _use_ more components _after_ integration.
```{r}
var_1 <- Stdev(panc_sub, reduction = "pca")^2
var_2 <- Stdev(panc.combined, reduction = "pca")^2

plot(var_1/sum(var_1))
points(var_2/sum(var_2), col="red")
```



Comparing counts, data, scale.data in Seurat object for given gene 

```{r}
g = "ACTB" # highly expressed across cell types
g = "CD14" # lowly expressed, cell-type specific
g = "HMGB2"
g = "GAPDH"
df_c <- GetAssayData(panc_sub, slot = "counts")[g,] %>% 
  data.frame() %>% rownames_to_column('cell') %>% mutate(slot="counts")
df_d <- GetAssayData(panc_sub, slot = "data")[g,] %>% 
  data.frame() %>% rownames_to_column('cell') %>% mutate(slot="data")
df_s <- GetAssayData(panc_sub, slot = "scale.data")[g,] %>% 
  data.frame() %>% rownames_to_column('cell') %>% mutate(slot="scale.data")

bind_rows(df_c, df_d, df_s) %>% 
  ggplot(aes(x=., fill=slot)) + 
  geom_histogram() + 
  facet_wrap(~slot, scales="free")
```

Signal, Noise and Signal-Noise
```{r}
counts = GetAssayData(panc_sub, slot = "counts")
mu = apply(counts, 1, mean)
std = apply(counts, 1, sd)

smoothScatter(log(mu), log(std))
smoothScatter(log(mu), log(mu/std))
smoothScatter(log(mu), mu/std)
smoothScatter(sqrt(mu), mu/std)
```




 

